---
title: "Linear Mixed Effect Models"
author: "Nick Freeland, Bernice Green, Gary Marmon"
date: "`r Sys.Date()`"
output: html_document
bibliography: ref.bib
nocite: '@*'
---

## Introduction 

<!-- #### What are LMMs?  -->
<!-- Include Key vocabulary -->
Linear Mixed-Effects Models can be used to model correlated data [@galecki2013] that often take the form of cross sectional data or longitudinal data. Cross sectional data feature individuals (level 1) nested in a geographical or social context (level 2), while longitudinal data feature individuals (level 2) measures over several occasions (level 1).[@bell2019fixed] Data can be nested in more than two levels based on the design of the study. 

Mixed-effects models are called “mixed” because they simultaneously model fixed and random effects, which account for clustering in a data set and the differeing relationships between & within clusters[@bell2019fixed]. Fixed effects model average trends, while random effects model the extent to which these trends vary across levels of some grouping factor [@brown2021introduction]. When the number of clusters is small but the number of observations per cluster is large, we can model that parameter as a fixed effect. Conversely, random effects may have a large number of clusters but a relatively small number of observations per cluster [@demidenko2013]. It is important to include the random effects in the model as fixed effects only give a partial picture of the heirarchial data as they do not reveal information about level 2 entities. Valuable information is lost about the relationship between the  [@bell2019fixed]

In the classical approach, all observations are assumed to be independent and identically distributed, but this assumption a=can lead to false results for clustered data. Observations between clustes are assumed independent but observations betweeen cluster are dependent as they belong to the same sub population.Mixed Effects Models can be seen as a combination of ANOVA, a fixed effects model, and VARCOMP, a random effects model.[@demidenko2013]

<!-- #### Why do we want to use LMMs? How are they beneficial? Fields used? Where are they most effectively applied? -->

Until more recently the only way to handle the type of data mixed-effects model does was through repeated measures ANOVAs. Mixed-effects models are much more versatile in handling variability within and across groups and can handle missing data, providing much better results than the ANOVAs. [@brown2021introduction]

@brown2021introduction gives a theoretical practical account of implementing mixed-effects models and their commonalities and differences with ANOVA, showing ANOVA cannot simultaneously take multiple sources of variation into account when observations are nested across participants, which lowers the ability to detect an effect. Mixed-effects modeling allows a researcher to examine the condition of interest while also considering variability within and across participants and items simultaneously and is a reasonable choice when ANOVA and multiple regression are not.[@brown2021introduction]

Linear mixed-models are most effective for clustered data with a hierarchical structure or repeated measures. They are also well-suited for time-series data, biological/medical data, and modeling shapes/images. [@demidenko2013]

The main application for mixed-effect models is in psychology due to the nature of their data and repeated observations across trial participants. However, the applications can extend into almost any field where the variability across a group/person is desired in the analysis. One such example is the use of mixed-effects models on published health data sets to explore the link between smoking and depression in which it was found “Smoking status is robustly associated with depression (or depressive symptomatology) at 1 ½ to 2 times the risk of nonsmoking across a variety of study designs, depression measurements, and participant populations” [@luger2014robust]. 

<!-- #### What are the limitations? -->

The critical slope of mixed-effects models id also oft discussed in literature, finding failure to include the critical slope in the test of an interaction can yield very high Type I error rates (Barr 2013). "When testing interactions in mixed designs with replications, it is critical to include the random slope corresponding to the highest-order combination of within-subject factors subsumed by each interaction of interest" (Barr 2013).

<!-- #### How does this apply to our project/data set? -->

By doing an extensive review of current literature our group aims to gain an understanding of Linear Mixed-Effect Models in order to create our own models and better our understanding. To do so we will use NFL play-by-play data from the nflverse package available in R. In our case we wish to find association between a NFL coach and play performance on the field using the nflverse data set, where the coaches are the random effects.   

## Methods

<!-- #### What are the assumptions of this model? -->

 The complex nature of mixed-effects models call into question the robustness of these models and brings more focus to the model assumptions. "Mixed-effects models involve complex fitting procedures and make several assumptions, in particular about the distribution of residual and random effects. Violations of these assumptions are common in real datasets, yet it is not always clear how much these violations matter to accurate and unbiased estimation" (Schielzeth et al. 2020).  Schielzeth et al. discuss the consequences of violations of these assumptions and the impact of missing random effect components on model estimates. The study found mixed-effects models to be very robust to violations of these assumptions, finding the estimates were unbiased (although imprecise) and missing random effect predictors had little effect on the fixed effect estimates but had systematic effects on the estimates of random effects (Schielzeth et al. 2020). 

Many technical papers have delved into the formulation and implementation of linear mixed-effects models, many following the lead of Bates et al. paper, "Fitting linear mixed-effects models using lme4" (2014). Bates et al. outlines the creation and implementation of the lme4 package as an extension of lmer function, which has become the predominant tool in the R language for fitting linear mixed-effect models. As for most model-fitting functions in R, the model is described in an lmer call by a formula, in this case including both fixed- and random-effects terms (Bates et al. 2014). The paper describes the structure of the model, the steps in evaluating the profiled deviance or REML criterion, and the structure of classes or types that represents such a model (Bates et al. 2014). One of the more controversial design decisions of lme4 has been to omit the output of p values (these can be found using parametric bootstrapping functionality). "While the null distributions (and the sampling distributions of non-null estimates) are asymptotically normal, these distributions are not t distributed for finite size samples — nor are the corresponding null distributions of differences in scaled deviances F distributed." (Bates et al. 2014). A common problem in mixed-model inference is the lack of analytical results for parameter estimates in complex situations including unbalanced or crossed designs. (Bates et al. 2014)


<!-- #### What is the formula for this model?  -->

Just as a linear model is described by the distribution of a vector-valued random response variable, Y, whose observed value is $y_{obs}$, a linear mixed model is described by the distribution of two vector-valued random variables: $Y$, the response, and $\beta$, the vector of random effects. In a linear model the distribution of Y is multivariate normal,

$$
Y ∼ N(Xβ + o, σ^2 W^{-1})
$$

Our basic model: $$epa = plays +  (1+plays|coach)$$

## Analysis and Results
### Data and Vizualisation

Our data featured a play by play analysis for every game in the 2021 season. We are attempting to model the coaches effect using the epa. We first decided to aggregate the data by game and team. We are assuming that all plays are rush or pass and will be using the pass percentage (pass_pct) to model the two different play options per game. After summarizing, we end up with the following table.

Variable            Meaning
-------------       -------------------------------------------------------------------
pbp.posteam         the team with possession of the ball (offense)
pop.posteam_type    specifies if the possessing team is home or away
pbp.game_id         the specific game id from the nfl
week                week number in the season that the game was played
season_type         flag that specifies if it is a regular (0) or post (1) season game
home_adv            flag that specifies home (0) or away(1)
coach               the coach of the team with possession (offensive plays)
opp_coach           the coach of the opposing team (defensive plays)
plays               total number of rush and pass plays given the team and game
pass_plays          number of pass plays given the team and game
pass_pct            the percentage of pass plays in the game calculated by pass_plays/plays
epa_per_play        the mean of all pass and rush plays given team and game

 
```{r message=FALSE, warning=FALSE, include=FALSE}
library(nflverse)
library(nflplotR)
library(tidyverse)
library(kableExtra)
```

NFL offense summaries from the 2021 season including coach and home/away splits.

```{r message=FALSE, warning=FALSE}
# loading play by play data from the 2021 NFL season
pbp <- nflreadr::load_pbp(2021)

# team summaries 
team_sum1 <- data.frame(pbp$game_id, pbp$home_coach, pbp$away_coach, pbp$posteam, pbp$posteam_type, pbp$pass, pbp$rush, pbp$epa, pbp$down, pbp$week, pbp$season_type)

team_sum2 <- team_sum1 %>%
  filter(pbp.rush == 1 | pbp.pass == 1, !is.na(pbp.down)) %>%
  group_by (pbp.posteam, pbp.posteam_type, pbp.game_id) %>%
  mutate(coach = ifelse(pbp.posteam_type == 'home',pbp.home_coach,pbp.away_coach)) %>%
  mutate(opp_coach = ifelse(pbp.posteam_type == 'away',pbp.home_coach,pbp.away_coach)) %>%
  mutate(home_adv = ifelse(pbp.posteam_type == 'home', 0, 1))%>%
  summarize(week = first(pbp.week),
            season_type = first(ifelse(pbp.season_type == 'REG', 0, 1)),
            home_adv = first(home_adv),
            coach = first(coach),
            opp_coach = first(opp_coach),
            plays = n(),
            pass_plays = sum(pbp.pass),
            pass_pct = pass_plays / plays,
            epa_per_play = round(mean(pbp.epa), digits = 2))


```

```{r, warning=FALSE, echo=T, message=FALSE, include=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(GLMsData)
library(ggfortify)
```


```{r, warning=FALSE, echo=TRUE}
summary(team_sum2)

ggplot(team_sum2, aes(x = ,
                     y = epa_per_play, 
                     color=coach)) +
  geom_boxplot() +
  labs(title = "")

ggplot(team_sum2, aes(x=pass_pct, y=epa_per_play, color=coach)) + geom_point()

ggplot(team_sum2,aes(x=week, y=epa_per_play, color=coach)) + geom_point()

ggplot(team_sum2, aes(x=plays, y=epa_per_play, color=coach)) + geom_point()


team_sum2$season_type <- as.factor(team_sum2$season_type)
team_sum2$home_adv <- as.factor(team_sum2$home_adv)

ggplot(team_sum2,aes(x=season_type, y=epa_per_play)) + geom_boxplot()

ggplot(team_sum2, aes(home_adv, epa_per_play)) + geom_boxplot()


```

```{r message=FALSE, warning=FALSE}
offense <- pbp %>%
  dplyr::group_by(team = posteam) %>%
  dplyr::summarise(off_epa = mean(epa, na.rm = TRUE))
defense <- pbp %>%
  dplyr::group_by(team = defteam) %>%
  dplyr::summarise(def_epa = mean(epa, na.rm = TRUE))
offense %>%
  dplyr::inner_join(defense, by = "team") %>%
  ggplot2::ggplot(aes(x = off_epa, y = def_epa)) +
  ggplot2::geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
  nflplotR::geom_mean_lines(aes(h_var = off_epa, v_var = def_epa)) +
  nflplotR::geom_nfl_logos(aes(team_abbr = team), width = 0.07, alpha = 0.7) +
  ggplot2::labs(
    x = "Offense EPA/play",
    y = "Defense EPA/play",
    title = "2021 NFL Offensive and Defensive EPA per Play"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 12, hjust = 0.5, face = "bold")
  ) +
  ggplot2::scale_y_reverse()
```


### Statistical Modeling

```{r message=FALSE, warning=FALSE, include=FALSE}
library(lme4)
```


```{r message=FALSE, warning=FALSE}
epa.lmer1 = lmer(epa_per_play ~ pass_pct + plays +
(1|coach) + (1|home_adv), data=team_sum2)
```

```{r}
epa.lmer1
```
- we observe the home advantage random effect approaches zero, concluding home advantage has a very small/no effect on the outcome of epa per play. 

```{r}
coef(epa.lmer1)
```
- each coach is assigned a different intercept, but the fixed effects are the same for all coaches. This model is called a Random Intercept model; we are accounting for baseline differences in epa per play. Alternatively, in a Random Slope model, each coach is allowed to have a different intercept, as well as different slopes for the effect of number of plays ran and percentage of pass plays. We implement a Random Slope model below:

```{r message=FALSE, warning=FALSE}
epa.lmer2 = lmer(epa_per_play ~ pass_pct + plays +
(1+pass_pct|coach) + (1+plays|coach), data=team_sum2)
```

```{r}
coef(epa.lmer2)
```
- here the model is expecting different baseline levels of plays ran and pass_pct

- despite the individual variation of pass_pct, all the values are negative and very close to each other. We see consistency with how often coaches throw the ball. The variation in number of plays ran is much wider, ranging from -7 to 6.

#### Testing for significance between models

```{r message=FALSE, warning=FALSE}
epa.lmer2.full = lmer(epa_per_play ~ pass_pct + plays +
                  (1+pass_pct|coach) + (1+plays|coach), data=team_sum2,
                  REML=FALSE)

epa.lmer2.null = lmer(epa_per_play ~ plays +
                  (1+plays|coach), data=team_sum2,
                  REML=FALSE)
```
```{r}
anova(epa.lmer2.full, epa.lmer2.null)
```
- we observe statistical significance between the models

```{r}
team_sum2f <- team_sum2 %>% filter(season_type==0)
epa.lmer2f = lmer(epa_per_play ~ pass_pct + plays +
(1+pass_pct|coach) + (1+plays|coach), data=team_sum2f)
intercepts2 <- rep(coef(epa.lmer2f)$coach[,1], each = 16)
slopes <- rep(coef(epa.lmer2f)$coach[,2], each = 16)
model_intercept <- as.numeric(fixef(epa.lmer2f)[1])
model_slope <- as.numeric(fixef(epa.lmer2f)[2])

ggplot(team_sum2f, aes(x = pass_pct, y = epa_per_play, color=coach)) + 
  geom_abline(slope = model_slope, intercept = model_intercept, 
              linetype = "solid", color = "black", size = 1) + 
  geom_abline(mapping = aes(slope = slopes, 
                            intercept = intercepts2, linetype = coach, color = coach), 
              linetype = "dashed", size = .4) +
  geom_point(aes(), size = 1) +
  theme(panel.background = element_blank(),         
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA),
        legend.position = "right", 
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 14)) +
  labs (x = "Pass Percent", y = "Epa Per Play")

slopes <- rep(coef(epa.lmer2f)$coach[,3], each = 16)
model_slope <- as.numeric(fixef(epa.lmer2f)[3])
ggplot(team_sum2f, aes(x = plays, y = epa_per_play, color=coach)) + 
  geom_abline(slope = model_slope, intercept = model_intercept, 
              linetype = "solid", color = "black", size = 1) + 
  geom_abline(mapping = aes(slope = slopes, 
                            intercept = intercepts2, linetype = coach, color = coach), 
              linetype = "dashed", size = .4) +
  geom_point(aes(), size = 1) +
  theme(panel.background = element_blank(),         
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA),
        legend.position = "right", 
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 14)) +
  labs (x = "Plays", y = "Epa Per Play")
```


```{r}
epa.lmer3 = lmer(epa_per_play ~  plays + home_adv +
                  (1+home_adv|coach) + (1+plays|coach), data=team_sum2,
                  REML=FALSE)
```
```{r}
epa.lmer3
```

```{r}
#attempt at visualizing the models  random intercepts plot was too messy

# team_sum2f <- team_sum2 %>% filter(season_type==0)
# epa.lmer1f = lmer(epa_per_play ~ pass_pct + plays +
# (1|coach) + (1|home_adv), data=team_sum2f)
# model_intercept <- as.numeric(fixef(epa.lmer1f)[1])
# model_slope <- as.numeric(fixef(epa.lmer1f)[3])
# intercepts <- rep(coef(epa.lmer1f)$coach[,1], each=16)
#
# ggplot(team_sum2f, aes(x = plays, y = epa_per_play)) +
#   geom_abline(slope = model_slope, intercept = model_intercept,
#               linetype = "solid", color = "black", size = 1) +
#   geom_abline(mapping = aes(slope = model_slope, intercept = intercepts),
#               linetype = "dashed", color = "grey70", size = 1) +
#   geom_point(aes(), size = 1, color = "grey70") +
#   geom_segment(aes(x = plays, xend = plays,
#   y = epa_per_play, yend = fitted(epa.lmer1f)),
#   color = "grey70") +
#   theme(panel.background = element_blank(),
#         panel.grid.major = element_blank(),
#         panel.grid.minor = element_blank(),
#         panel.border = element_rect(colour = "black", fill = NA),
#         legend.position = "none",
#         axis.text = element_text(size = 14),
#         axis.title = element_text(size = 14)) +
#   labs (x = "Play", y = "Epa Per Play")
```

```{r}
epa.lmer3f = lmer(epa_per_play ~  plays + home_adv +
                  (1+home_adv|coach) + (1+plays|coach), data=team_sum2f,
                  REML=FALSE)
intercepts2 <- rep(coef(epa.lmer3f)$coach[,1], each = 16)
slopes <- rep(coef(epa.lmer3f)$coach[,2], each = 16)
model_intercept <- as.numeric(fixef(epa.lmer3f)[1])
model_slope <- as.numeric(fixef(epa.lmer3f)[2])

ggplot(team_sum2f, aes(x = plays, y = epa_per_play, color=home_adv)) + 
  geom_abline(slope = model_slope, intercept = model_intercept, 
              linetype = "solid", color = "black", size = 1) + 
  geom_abline(mapping = aes(slope = slopes, 
                            intercept = intercepts2, linetype = coach), 
              linetype = "dashed", size = .4, color="grey70") +
  geom_point(aes(color=home_adv), size = 1, color = "grey70") + 
  geom_segment(aes(x = plays, xend = plays, 
                   y = epa_per_play, yend = fitted(epa.lmer3f)), 
               ) +
  theme(panel.background = element_blank(),         
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA),
        legend.position = "right", 
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 14)) +
  labs (x = "Plays", y = "Epa Per Play")
```


### Conlusion

## References
```{r}
```


